---
title: "Exploring my Pocket archive"
output:
  html_document:
    keep_md: yes
date: "January, 2015"
---

Im am using [Pocket](http://getpocket.com) as my main bookmarking service to store articles I find on the web during the day and want to read later. After I used it constantly for more than half a year, I wanted to evaluate how I actually use Pocket and gain some insights into what kind of articles are of interest to me. The file pocket_data.rmd in this Github-Repository contains all necessary calls, in case you want to use it as a starting point for your own analysis. Required packages are XML, dplyr, lubridate and ggplot2.

### Retrieving the data

Pocket offers [export functionality](https://getpocket.com/export) as an HTML-file and by using an API. Since I do not want to perform the analysis on a daily basis, a simple manual download seems more appropriate. The resulting file is used as baseline for this analysis.
```{r include=FALSE}
require(XML)
require(dplyr)
require(lubridate)
require(ggplot2)
filePath <- 'C:/Users/florian.mueller/Dropbox/data_analysis/pocket_data/20150110_ril_export.html'
```
As a first step, the weblinks of stored articles are retrieved by searching for \<a\> HTML elements and in particular their attributes:
```{r}
doc.html = htmlTreeParse(filePath, useInternal = TRUE)
doc.text = xpathApply(doc.html, '//a', xmlAttrs)
names(unlist(doc.text[1]))
```
We get three attributes:

* href: The link
* time_added: The timestamp, when the link was added to pocket
* tags: Tags assigned to a link

Since I do not use tags, I will concentrate on the first two attributes for the analysis. The provided link offers some potential for direct feature extraction. For example, the base domain is interesting in order to identify common sources for articles from the web.

The base domains can be retrieved by using a Regular Expression on the list of links. First the href-attributes are selected (each third row) and afterwards matched:
```{r}
doc.flatlist <- unlist(doc.text)
doc.length <- length(doc.flatlist)
doc.rawLinks <- doc.flatlist[seq(1,924,3)]

matches <- regexpr("([^\\.|/]+)\\.([a-z]+)/", doc.rawLinks, perl=TRUE)
doc.domains <- regmatches(doc.rawLinks, matches)
head(doc.domains)
```
The time attributes counts seconds, starting from 1970-01-01 and can be transformed into readable dates by using POSIX date formats:
```{r}
doc.rawTimeAdded <- doc.flatlist[seq(2,924,3)]
doc.parsedTimeAdded <- lapply(lapply(doc.rawTimeAdded, as.integer), 
                              function(x) as.POSIXct(x, origin="1970-01-01",tz = "GMT"))
```
For easier access, several time related features are extracted. All features are then stored in a single data frame:
```{r}
doc.df <- data.frame(doc.rawLinks,
                     as.integer(doc.rawTimeAdded),
                     doc.domains,
                     sapply(doc.parsedTimeAdded, year), 
                     sapply(doc.parsedTimeAdded, month),
                     sapply(doc.parsedTimeAdded, week),
                     sapply(doc.parsedTimeAdded, day),
                     sapply(doc.parsedTimeAdded, wday),
                     sapply(doc.parsedTimeAdded, hour))

names(doc.df) <- c("Link", "Added.on", "Domain", "Year","Month","Week","Day","Weekday","Hour")
```

### Analysis

Frist of all, let's see how pocket was used by me. To this end, the number of articles are plotted, based on when they were added:

```{r, echo=FALSE}
ana.noOfArticles <- arrange(doc.df, Added.on) %>% 
  mutate(Date = as.POSIXct(Added.on, origin="1970-01-01",tz = "GMT")) %>% 
  select(Date) %>% 
  cbind(1:nrow(doc.df))

names(ana.noOfArticles) <- c("Date", "ArticleNumber")

ggplot(ana.noOfArticles, aes(x=Date, y=ArticleNumber)) +
  ggtitle("Archive size during usage") +
  geom_point(shape=1) 
```

I started using Pocket in 2012 in a rather limited way, stopping at the end of 2012 at 24 stored articles. This stop continued until mid of 2014. At first, just a brief usage is visible. Then, the usage skyrocketed, resulting in a total number of 308 articles. One possible explanation could be that the mode of usage changed significantly comparing to the first usage in 2012, since the number of added articles per month apprears to be much larger:

```{r, echo=FALSE}
mutate(doc.df, 
       curr_month = as.POSIXct(years(Year) + months(Month) - months(1), 
                               origin = "00-01-01", zt = "GMT")) %>% 
  group_by(curr_month) %>% summarize(number = n()) %>% 
  ggplot(aes(x=curr_month, y=number)) + geom_bar(stat="identity") +
  ggtitle("Number of articles per month")
grouped_by_Year <- group_by(doc.df, Year, Month)
```

After stopping to use Pocket at the end of 2012, I gave it a shot again in April 2014. However, the real start appreas to be in May. In fact, after starting to use pocket again, the usage in the second to fifth month (May 2014 to August 2014) is very high, followed by a reduced usage in the next four month (September 2014 to December 2014).

In order to check, if the mode of usage changed, the most visited domains in 2012 and 2014 are compared:

```{r, echo=FALSE}
grouped_by_Year_Domain <- group_by(doc.df, Year, Domain)
summarize(grouped_by_Year_Domain, number = n()) %>% 
  arrange(desc(number)) %>% 
  filter(Year == 2012, number > 1) %>% head(10) %>% 
  ggplot(aes(x=Domain, y=number)) + 
  geom_bar(stat="identity") + 
  ggtitle("Top domains 2012") +
  theme(axis.text.x = element_text(angle=45, vjust=0.5, size=12))
```

There are only four domains in 2012 that contributed more than one article to my Pocket archive. The top domain is [Lifehacker](http://lifehacker.com), which incidentally lead me to use Pocket in the first place. Second place goes to [youtube](http://youtube.com), followed by [ted.com](http://ted.com) as another video-resource and [Süddeutsche](http://sueddeutsche.de), a German newspaper.

```{r, echo=FALSE}
summarize(grouped_by_Year_Domain, number = n()) %>% 
  arrange(desc(number)) %>% 
  filter(Year == 2014, number > 1) %>% head(10) %>% 
  ggplot(aes(x=Domain, y=number)) + 
  geom_bar(stat="identity") + 
  ggtitle("Top domains 2014") +
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1, size=12))
```

In 2014 the variety of domains has changed, allowing a display of top 10 domains with more than one article contribution. The top contributor is now [Wired](wired.com). [Lifehacker](http://lifehacker.com) and [youtube](http://youtube.com) are still strong, followed by two newspapers, [Spiegel](http://spiegel.de) and [New York Times](http://nytimes.com). In order to discover whether the top domains constantly contribute to the archive or whether there are certain months with a lot of consumption of a particular domain, the number of articles per month are plotted:

```{r, echo=FALSE}
ana.baseDf <- mutate(doc.df, 
                     curr_month = as.POSIXct(years(Year) + months(Month) - months(1), 
                                             origin = "00-01-01", zt = "GMT"))
ana.baseDf %>% group_by(Domain, curr_month) %>% summarize(number = n()) %>%
  filter(Domain == "wired.com/" | Domain == "lifehacker.com/" | Domain == "sueddeutsche.de/" |
           Domain == "ted.com/" | Domain == "youtube.com/" | Domain == "spiegel.de/" |
           Domain == "nytimes.com/", year(curr_month) == 2014) %>% 
  ggplot(aes(x=curr_month, y=number, colour = Domain)) + geom_line(size = 1.5) +
  ggtitle("Articles per month of top domains in 2014")
```

Most of the wired.com articles were added in May to August, afterwards the number declines. Lifehacker articles are strong in May but decline afterwards rather quickly. Youtube appears to be somehow constant, since it does not display a decline but more of an on/off pattern. Let's see how much the identified top domains contribute to the total number of articles per month:

```{r, echo=FALSE}
ana.baseDf %>% 
  mutate(inTopList = Domain %in% c("wired.com/", "lifehacker.com/", 
                                   "sueddeutsche.de/", "ted.com/", "youtube.com/", 
                                   "spiegel.de/", "nytimes.com/")) %>%
  filter(Year == 2014) %>%
  group_by(curr_month) %>%
  summarize(proportionInTopList = mean(inTopList)) %>%
  arrange(desc(proportionInTopList)) %>%
  ggplot(aes(x=curr_month, y=proportionInTopList)) +
  geom_bar(stat="identity") +
  ggtitle("Proportion of top domains per month in 2014")
```

After July, a notable decline in share can be noted, even though the share recovers in the following months until December, where the lowest share so far is achieved. This may indicate a shift in interests or a greater variety of sources. Thus, the proportion of unique domains with only one article contributing to a month is checked as a next step.

```{r, echo=FALSE}
ana.baseDf %>% 
  group_by(Domain, curr_month) %>% 
  summarize(number = n()) %>%
  mutate(one_or_less = number == 1) %>%
  group_by(curr_month) %>%
  summarize(new_number = mean(one_or_less)) %>%
  ggplot(aes(x=curr_month, y=new_number)) +
  ggtitle("Proportion of domains with only one article per month") +
  geom_bar(stat="identity")
```

It appears as if the proportion of single-article domains is rising, indicating a larger variety of sources. This observation is also backed by the following query, checking the number of articles per domain per month. Starting in July 2014, the number average number of articles per domain decreases, getting quite close to one article per domain in December 2014.

```{r, echo=FALSE}
ana.baseDf %>% 
  group_by(curr_month) %>%   
  summarize(number = length(unique(Domain)), total = n()) %>%
  mutate(number_of_articles = total  / number) %>%
  ggplot(aes(x=curr_month, y=number_of_articles)) +
  ggtitle("Average number of articles per domain") +
  geom_bar(stat="identity")
```

```{r, echo=FALSE}
ana.baseDf %>% 
  group_by(Weekday) %>%   
  summarize(number = n()) %>%
  ggplot(aes(x=Weekday, y=number)) + 
  geom_bar(stat="identity") + 
  ggtitle("Number of articles added per weekday")
```

```{r, echo=FALSE}
ana.baseDf %>% 
  group_by(Hour) %>%   
  summarize(number = n()) %>%
  ggplot(aes(x=Hour, y=number)) + 
  geom_bar(stat="identity") + 
  ggtitle("Number of articles added per hour")
```

